The mid-term impact of COVID-19 on human mobility patterns in Latin American countries\footnote{\textbf{Citation}: Cabrera-Arnau, C., González-Leonardo, M., Nasuto, A., Neville, R., Rowe, F. (2023). The mid-term impact of COVID-19 on human mobility patterns in Latin American countries}

### 2024-01-17 manuscript-3-old/manuscript-latin-countries.tex
compiling-manuscript

The COVID-19 pandemic globally resulted in disruptive shocks to the
human mobility system. Governments stringency measures and the economic
downturn caused by the pandemic constrained international migration
({``International Migration 2020: Highlights''} 2021; González-Leonardo
and Spijker 2022; González-Leonardo et al. 2023) and population
movements within countries (Nouvellet et al. 2021; González-Leonardo,
Rowe, and Fresolone-Caparrós 2022; Wang et al. 2022; Rowe,
González-Leonardo, and Champion 2023). During early stages of the
pandemic, big cities became the main epicentres of COVID-19 infections
and deaths (Pomeroy and Chiney 2021) due to their higher connectivity
via air travel, higher population density and concentration of
public-facing jobs (Brandén et al. 2020; Bhadra, Mukherjee, and Sarkar
2020). As a result, governments implemented non-pharmaceutical
interventions, such as lockdowns and business closures, leading to
increases in unemployment and removing the urban vibrancy (Blustein et
al. 2020; BBC news 2023). In addition, remote work and online teaching
reduced the need for commuting, living close to work and education
centers, and added pressure for individuals living in small and crowded
spaces to move out of cities and look for more affordable and
comfortable housing (Nathan and Overman 2020; US Bureau of Labor
Statistics 2021). These changes are believed to have reduced the
attractiveness of cities during the pandemic, while suburbs and rural
areas have gained popularity (Florida, Rodríguez-Pose, and Storper
2021). As a result, a number of headlines have claimed the emergence of
an `urban exodus' to areas with lower population densities (e.g. (The
Guardian 2020)).

Despite their potential rise in popularity, remote locations offer
limited availability and diversity of job opportunities, and tend to
lack infrastructure and services, such as schooling and health, to
effectively accommodate a large population influx (Pinilla and Sáez
2021; \emph{OECD Regions and Cities at a Glance}, n.d.). They generally
lack high-speed Internet connection which is vital for teleworking (Chen
and Wellman 2004). Additionally, most jobs cannot be done remotely
{``OECD Social, Employment and Migration Working Papers''} (n.d.), and
some companies have returned to in-person office work or have
implemented hybrid forms of work, following the relaxation of stringency
measures (McKinsey \& Company 2021). Schools and universities have also
returned to face-to-face teaching. Therefore, living close to work or
study places remains relevant. Businesses and leisure activities in
cities have gradually returned to normal, reinvigorating the economic
functioning and urban life. Collectively, these arguments suggest that a
potential increase in counterurbanisation movements during the pandemic
was concentrated during periods of stringency measures and is likely to
be short term. Big cities have successfully weathered previous pandemics
(Huremović 2019; Glaeser 2020) and are likely to remain attractive
places to live. Thanks to agglomeration economies, cities are normally
regarded as centres for economic growth, clusters of talent, consumer
bases and spaces for face‐to‐face interaction and diversity (Storper and
Venables 2004; Florida, Rodríguez-Pose, and Storper 2021).

Recent empirical evidence has shown a decline on human mobility during
periods of severe stringency measures, coupled with a rise of movements
from cities to less populated areas and a slowdown of urbanisation
movements across some countries of the Global North: the United States
(Ramani and Bloom 2021), United Kingdom (Rowe, González-Leonardo, and
Champion 2023; Wang et al. 2022), Spain (González-Leonardo et al. 2022;
González-Leonardo, Rowe, and Fresolone-Caparrós 2022), Germany {[}@;
Stawarz et al. (2022){]}, Sweden (Vogiazides and Kawalerowicz 2022),
Norway (Tønnessen 2021), Australia (Perales and Bernard 2022) and Japan
Kotsubo and Nakaya (2022). The majority of movements away from cities
headed to their suburbs (Vogiazides and Kawalerowicz 2022; Stawarz et
al. 2022) or rural areas in their close proximity (González-Leonardo et
al. 2022; González-Leonardo, Rowe, and Fresolone-Caparrós 2022) showing
in some instances a ``doghnut effect'', with a decline in population
movements directed to cities and a rise in the inflow of people to less
dense surrounding areas (Ramani and Bloom 2021). Studies demonstrated
that professionals (Tønnessen 2021) and individuals with a high income
(Haslag and Weagley 2021), those who are potentially able to practice
remote working, underpinned movements away from large cities. In
addition, research has shown that a large share of counterurbanisation
movements is headed to touristic locations with a high concentration of
second homes, suggesting that middle and high class individuals may have
been key actors in movements from cities to less populated areas during
the pandemic (González-Leonardo, Rowe, and Fresolone-Caparrós 2022).
However, most studies suggested that variations in internal population
movements during the pandemic are likely to be temporary and have not
altered previous macro-structures of human mobility across the
rural-urban hierarchy (Rowe, González-Leonardo, and Champion 2023).

In Latin America and the Caribbean, a reduction in internal population
movements was documented during periods of high stringency (Aromí et al.
2023), but there is no evidence on how human mobility has changed across
the rural-urban continuum. Anecdotal reports have shown that movements
away from cities seem to have accelerated in two countries of the Global
South, India (Irudaya Rajan, Sivakumar, and Srinivasan 2020) and South
Africa (Ginsburg et al. 2022), while urbanisation trends have
decelerated. However, lack of suitable up-to-date data has prevented us
from properly assessing the ``urban exodus'' hypothesis in Global South
countries, as well as the extent, spatial patterns, and durability of
potential changes on internal population movements. The evidence for
these two countries suggested that increasing counterurbanisation
movements during the pandemic in India and South Africa could be
underpinned by internal migrants residing in cities who returned to
their hometowns due to a rise in unemployment (Irudaya Rajan, Sivakumar,
and Srinivasan 2020; Ginsburg et al. 2022). However, recent studies have
shown that urban residents from wealthy neighborhoods were more likely
to move to rural areas than those from low income neighborhoods during
the first wave of the pandemic in Brazil, Colombia, Mexico, Indonesia,
Philippines and South Africa (Lucchini et al. 2023).

It is known that human mobility within countries tends to vary in
systematic ways with development levels (Bell et al. 2015). In Latin
America and the Caribbean, 28.6\% of the population was living on less
than \$6.8 per day in 2020 (World Bank" 2023) and the informality rate
was about 53\% in 2016 (Salazar-Xirinachs 2018). Thus, we could expect
different pandemic impacts on the patterns of internal population
movements between countries in the Global North and the Global South.
The way in which COVID-19 unfolded in each country, with varying levels
of stringency measures, may have also produced different outcomes on
human mobility.

Traditionally, censuses have been used to analyse human mobility
patterns within Global South countries, which lack register data (Bell
et al. 2015; Bernard et al. 2017). Censuses, however, are normally
updated every 10 years, lacking the temporal granularity to explore
population movements over yearly, monthly or short-time intervals.

Digital traces data from mobile phone applications, on the other hand,
provide a unique opportunity to capture internal population movements
with small spatial and temporal granularity in multiple countries
(Green, Pollock, and Rowe 2021; Rowe 2021). Drawing on geographically
granular Meta-Facebook data from March 2020 to March 2022, we aim to
analyse how the pandemic has changed the patterns of human mobility
across the rural-urban hierarchy in four Latin American countries:
Argentina, Chile, Colombia and Mexico. We aim to address the following
research questions: 1) To what extent did people leave cities and move
to less populated areas in these countries? 2) Which are the spatial
patterns of potential changes on internal population movements? 3) How
have the patterns of human mobility changed over the course of the
pandemic? 4) Are potential changes likely to endure the pandemic?

@@

The COVID-19 pandemic led to an overall decline in human mobility within
countries (Nouvellet et al. 2021). Research has shown that internal
population movements across the rural-urban continuum decreased in the
United States (Ramani and Bloom 2021), several European countries, Japan
and Austria during 2020 (Rowe, González-Leonardo, and Champion 2023).
Declines ranged from 2.5\% in Spain to 8.5\% in Australia. The largest
drops occurred during periods of national lockdowns and the
implementation of other stringency measures, such as mobility
restrictions, work and school closing, while levels of internal
population movements recovered or exceeded pre-pandemic figures when
lockdowns were lifted. Declines were partially attributed to a reduction
in involuntary migration during lockdowns, a loss of labour market
dynamism due to the pandemic, fewer people changing jobs and increasing
remote work (Perales and Bernard 2022).

COVID-19 has also altered patterns of internal population movements
across the rural-urban hierarchy (Rowe et al. 2022). In large metro
areas across the United States, population movements, as well as
households and businesses, shifted from central cities to suburbs and
exurbs during 2020 (Ramani and Bloom 2021). These authors label this
trend as a ``doughnut effect'', reflecting a decrease in population
inflows and activities in city centers and a growth in suburban rings.
They found a sizeable ``doghnut effect'' in large cities, a smaller
effect in medium-sized cities and almost no effect for small cities. A
similar trend was observed in Norway (Tønnessen 2021), Germany (Stawarz
et al. 2022), Sweden (Vogiazides and Kawalerowicz 2022) and Japan
(Fielding and Ishikawa 2021; Kotsubo and Nakaya 2022), where net
migration rates in big cities declined, while population movement to
their suburbs increased. In addition, this trend was accompanied by a
deceleration of urbanisation trends and unusual population gains in
rural areas due to an acceleration in counterurbanisation movements,
mostly in rural locations which are known to be holiday destinations
with a high concentration of second homes and are generally located in
the close proximity to big cities. An increase of counterurbanisation
was the main pandemic outcome on human mobility in Spain
(González-Leonardo et al. 2022; González-Leonardo, Rowe, and
Fresolone-Caparrós 2022), the United Kingdom (Wang et al. 2022; Rowe et
al. 2022) and Australia (Perales and Bernard 2022), where there is no
evidence of a ``doghnut effect'', since substantial variations in
suburbs were not found across these countries.

However, studies suggest that variations are likely to be temporary
(Rowe, González-Leonardo, and Champion 2023). In Spain, the urbanisation
movement returned to pre-pandemic levels when the lockdown ended in
mid-June 2020 (González-Leonardo et al. 2022) and unusually high levels
of counterurbanisation persisted over 2021, although the majority of
movements continued to occur between and within urban areas
(González-Leonardo, Rowe, and Fresolone-Caparrós 2022). In Australia,
the pandemic has caused minor changes in spatial patterns of internal
migration, and its effects were minimal by the end of 2020 (Perales and
Bernard 2022). In the United Kingdom, mobility patterns returned to
those registered prior to the pandemic after the easing of
non-pharmaceutical interventions (Rowe et al. 2022; Wang et al. 2022).
These findings suggest that COVID-19 generated shock waves leading to
temporary changes in the patterns of internal population movement across
the rural-urban continuum, but it has not significantly altered the
prevalent macro-structures of population movement within countries.

Despite some evidence in the Global North, lees is known about the
impact of COVID-19 on human mobility in the Global South countries. In
Latin America and the Caribbean, internal population movements declined
by 10\% during periods of severe stringent measures, from 16-19\% in
Bolivia, Ecuador and Argentina to less than 3\% in Paraguay and
Venezuela (Aromí et al. 2023). However, there is no evidence about
pandemic outcomes on human mobility across the rural-urban continuum. To
date, surveys carried out in India (Irudaya Rajan, Sivakumar, and
Srinivasan 2020) and South Africa (Ginsburg et al. 2022) have shown
anecdotal evidence suggesting that urbanisation trends declined during
the pandemic, while movements from cities to rural areas increased. Both
studies pointed out that a number of rural residents initiating a
migration decreased during 2020 and some labourers in major cities seem
to have returned to their hometowns due to job losses, since stringency
measures have extensively impacted the functioning of the economy
(Ghosh, Seth, and Tiwary 2020). Many internal migrants residing in
Indian cities lost their jobs and were unable to afford basic
necessities (Nair and Verma 2020). Collectively, these findings suggest
that vulnerable populations seem to have played a role in increasing
counterurbanisation movements in Global South countries. However, recent
work has shown that high-wealth groups residing in cities across Mexico,
Colombia, Brazil, Indonesia, Philippines and South Africa were on
average 159\% more likely to have moved to rural area compared to those
of low-wealth groups during early stages of the pandemic (Lucchini et
al. 2023), in line with the evidencee found for Global North countries.

@@

Until 1980, rural to urban migration dominated internal population
movements in Latin America with significant levels of population
redistribution, especially during the rapid industrialisation process
between the 1950s and 1970s when a critical mass of individuals moved
from villages and towns to urban areas (Firebaugh 1979; A. Lattes 1995;
J. Sobrino 2012). Urban growth due to internal mobility was primarily
concentrated in large urban centres resulting in a significant
population imbalance between the chief cities and other settlements
(Pinto da Cunha 2002; A. E. Lattes, Rodríguez, and Villa 2017). As a
consequence, Latin America shows high rates of urbanisation with more
than 81\% of the population living in cities, the highest figure after
North America, and about 46\% of urban residents settled in cities with
more than 1 million inhabitants (Nations" 2019).

Currently, internal population movements amongst cities dominate the
migratory system in Latin America, while rural to urban flows are of
less importance (Bernard et al. 2017; Rodríguez-Vignoli and Rowe 2018;
Nations" 2019). According to the 2010 census round, 80\% of internal
migrants moved between cities (Rodríguez-Vignoli and Rowe 2018).
Contrary to the period preceding the 1980s, medium-sizes cities (those
between 500K and 1 million inhabitants) experienced the highest net
migration rates, while large cities (\textgreater1 million) presented
balanced rates and small cities (\textgreater500K) lost individuals by
internal population movements (Rodríguez-Vignoli and Rowe 2018).
Changing patterns in human mobility within Latin American countries were
driven by a decline in internal migration since the 1980s crisis (Chávez
Galindo et al. 2016) and deconcentration dynamics in large urban
centres, such as Mexico City (Jaime Sobrino 2006) or Santiago de Chile
(González Ollino and Rodríguez Vignoli 2006), where long distance
inflows have declined over time. At the same time, there has been an
increasing domestic and foreign investment in export-oriented activities
or tourism industries in certain middle-sized cities, fostering the
geographic dispersal of employment opportunities and increasing the
attractiveness of these cities (Brea 2003; Pérez-Campuzano 2013; Chávez
Galindo et al. 2016).

Suburbanisation also represents an important type of internal population
movement and has increased over time due to the growing expansion of
Latin American cities (Graizbord and Acuña 2007; Chávez Galindo et al.
2016). These movements are underpinned by flows of middle- and
upper-class families moving away from central cities to surrounding
auto-segregated areas of wealthy individuals (Borsdorf 2003; Rodríguez
Vignoli and Rowe 2017). Low-income populations also settle in specific
areas within suburbs where the cost of living is usually cheaper than in
cities (Janoschka 2002; Rodríguez Vignoli and Rowe 2017). A large
proportion of both wealthy individuals and low-income residents commute
daily to central cities, mainly for work reasons (Chávez Galindo et al.
2016). Recently, reurbanisation dynamics have also been observed due to
gentrification processes in specific central areas (J. Sobrino 2012;
Chávez Galindo et al. 2016). In this paper, we analyse how the COVID-19
pandemic has changed patterns of internal population movements across
the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and
Argentina.

@@

\begin{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-aromuxed2023}{}}%
Aromí, Daniel, María Paula Bonel, Julian Cristia, Martín Llada, Juan
Pereira, Xiomara Pulido, and Julieth Santamaria. 2023. {``\#StayAtHome:
Social Distancing Policies and Mobility in Latin America and the
Caribbean.''} \emph{Economía} 22 (1).
\url{https://doi.org/10.31389/eco.4}.

\leavevmode\vadjust pre{\hypertarget{ref-BBCnews2023}{}}%
BBC news. 2023. {``Unemployment: Who Are the Millions of Britons Not
Working?''} \url{https://www.bbc.co.uk/news/business-52660591}.

\leavevmode\vadjust pre{\hypertarget{ref-bell2015}{}}%
Bell, Martin, Elin Charles-Edwards, Philipp Ueffing, John Stillwell,
Marek Kupiszewski, and Dorota Kupiszewska. 2015. {``Internal Migration
and Development: Comparing Migration Intensities Around the World.''}
\emph{Population and Development Review} 41 (1): 33--58.
\url{https://doi.org/10.1111/j.1728-4457.2015.00025.x}.

\leavevmode\vadjust pre{\hypertarget{ref-bernard2017}{}}%
Bernard, Aude, Francisco Rowe, Martin Bell, Philipp Ueffing, and Elin
Charles-Edwards. 2017. {``Comparing Internal Migration Across the
Countries of Latin America: A Multidimensional Approach.''} Edited by
Osman Alimamy Sankoh. \emph{PLOS ONE} 12 (3): e0173895.
\url{https://doi.org/10.1371/journal.pone.0173895}.

\leavevmode\vadjust pre{\hypertarget{ref-bhadra2020}{}}%
Bhadra, Arunava, Arindam Mukherjee, and Kabita Sarkar. 2020. {``Impact
of Population Density on Covid-19 Infected and Mortality Rate in
India.''} \emph{Modeling Earth Systems and Environment} 7 (1): 623--29.
\url{https://doi.org/10.1007/s40808-020-00984-7}.

\leavevmode\vadjust pre{\hypertarget{ref-blustein2020}{}}%
Blustein, David L., Ryan Duffy, Joaquim A. Ferreira, Valerie
Cohen-Scali, Rachel Gali Cinamon, and Blake A. Allan. 2020.
{``Unemployment in the Time of COVID-19: A Research Agenda.''}
\emph{Journal of Vocational Behavior} 119 (June): 103436.
\url{https://doi.org/10.1016/j.jvb.2020.103436}.

\leavevmode\vadjust pre{\hypertarget{ref-borsdorf2003}{}}%
Borsdorf, Axel. 2003. {``Cómo Modelar El Desarrollo y La Dinámica de La
Ciudad Latinoamericana.''} \emph{EURE (Santiago)} 29 (86).
\url{https://doi.org/10.4067/s0250-71612003008600002}.

\leavevmode\vadjust pre{\hypertarget{ref-branduxe9n2020}{}}%
Brandén, Maria, Siddartha Aradhya, Martin Kolk, Juho Härkönen, Sven
Drefahl, Bo Malmberg, Mikael Rostila, Agneta Cederström, Gunnar
Andersson, and Eleonora Mussino. 2020. {``Residential Context and
COVID-19 Mortality Among Adults Aged 70 Years and Older in Stockholm: A
Population-Based, Observational Study Using Individual-Level Data.''}
\emph{The Lancet Healthy Longevity} 1 (2): e80--88.
\url{https://doi.org/10.1016/s2666-7568(20)30016-7}.

\leavevmode\vadjust pre{\hypertarget{ref-Brea03}{}}%
Brea, J. 2003. {``Population Dynamics in Latin America.''}
\emph{Population Bulletin} 58: 3--36.
\url{https://population.un.org/wup/Publications/Files/WUP2018-Report.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-chuxe1vezgalindo2016}{}}%
Chávez Galindo, Ana María, Jorge Rodríguez Vignoli, Mario Acuña, Jorge
Barquero, Daniel Macadar, José Marcos Pinto da Cunha, and Jaime Sobrino.
2016. {``Migración Interna y Cambios Metropolitanos.''} \emph{Revista
Latinoamericana de Población} 10 (18): 7--41.
\url{https://doi.org/10.31406/relap2016.v10.i1.n18.1}.

\leavevmode\vadjust pre{\hypertarget{ref-Chen2004}{}}%
Chen, Wenhong, and Barry Wellman. 2004. {``The Global Digital Divide -
Within and Between Countries.''} \emph{IT \& Society} 1: 39--45.
\url{http://www.ifs.tuwien.ac.at/~dieter/teaching/GmA/Chen2004.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-fielding2021}{}}%
Fielding, Tony, and Yoshitaka Ishikawa. 2021. {``COVID{-}19 and
Migration: A Research Note on the Effects of COVID{-}19 on Internal
Migration Rates and Patterns in Japan.''} \emph{Population, Space and
Place} 27 (6). \url{https://doi.org/10.1002/psp.2499}.

>>>>

The COVID-19 pandemic globally resulted in disruptive shocks to the
human mobility system. Governments stringency measures and the economic
downturn caused by the pandemic constrained international migration
(\textbf{internat2021?}; \textbf{gonzález-leonardo2022?};
\textbf{gonzález-leonardo2023?}) and population movements within
countries (\textbf{nouvellet2021?}; \textbf{gonzález-leonardo2022a?};
Wang et al. 2022; \textbf{rowe2023?}). During early stages of the
pandemic, big cities became the main epicentres of COVID-19 infections
and deaths (\textbf{Pomeroy2021?}) due to their higher connectivity via
air travel, higher population density and concentration of public-facing
jobs (\textbf{brandén2020?}; \textbf{bhadra2020?}). As a result,
governments implemented non-pharmaceutical interventions, such as
lockdowns and business closures, leading to increases in unemployment
and removing the urban vibrancy (\textbf{blustein2020?};
\textbf{BBCnews2023?}). In addition, remote work and online teaching
reduced the need for commuting, living close to work and education
centers, and added pressure for individuals living in small and crowded
spaces to move out of cities and look for more affordable and
comfortable housing (\textbf{nathan2020?}; \textbf{USBureauLabor2021?}).
These changes are believed to have reduced the attractiveness of cities
during the pandemic, while suburbs and rural areas have gained
popularity (\textbf{florida2021?}). As a result, a number of headlines
have claimed the emergence of an `urban exodus' to areas with lower
population densities (e.g. (\textbf{TheGuardian2020?})).

Despite their potential rise in popularity, remote locations offer
limited availability and diversity of job opportunities, and tend to
lack infrastructure and services, such as schooling and health, to
effectively accommodate a large population influx
(\textbf{Pinilla2021?}; \textbf{oecdreg?}). They generally lack
high-speed Internet connection which is vital for teleworking
(\textbf{Chen2004?}). Additionally, most jobs cannot be done remotely
(\textbf{oecdsoc?}), and some companies have returned to in-person
office work or have implemented hybrid forms of work, following the
relaxation of stringency measures (\textbf{McKinsey21?}). Schools and
universities have also returned to face-to-face teaching. Therefore,
living close to work or study places remains relevant. Businesses and
leisure activities in cities have gradually returned to normal,
reinvigorating the economic functioning and urban life. Collectively,
these arguments suggest that a potential increase in counterurbanisation
movements during the pandemic was concentrated during periods of
stringency measures and is likely to be short term. Big cities have
successfully weathered previous pandemics (\textbf{huremovic2019?};
\textbf{Glaeser20?}) and are likely to remain attractive places to live.
Thanks to agglomeration economies, cities are normally regarded as
centres for economic growth, clusters of talent, consumer bases and
spaces for face‐to‐face interaction and diversity
(\textbf{storper2004?}; \textbf{florida2021?}).

Recent empirical evidence has shown a decline on human mobility during
periods of severe stringency measures, coupled with a rise of movements
from cities to less populated areas and a slowdown of urbanisation
movements across some countries of the Global North: the United States
(\textbf{ramani2021?}), United Kingdom (\textbf{rowe2023?}; Wang et al.
2022), Spain (\textbf{gonzález-leonardo2022b?};
\textbf{gonzález-leonardo2022a?}), Germany {[}@;
(\textbf{stawarz2022?}){]}, Sweden (\textbf{vogiazides2022?}), Norway
(\textbf{tønnessen2021?}), Australia (\textbf{perales2022?}) and Japan
(\textbf{kotsubo2022?}). The majority of movements away from cities
headed to their suburbs (\textbf{vogiazides2022?};
\textbf{stawarz2022?}) or rural areas in their close proximity
(\textbf{gonzález-leonardo2022b?}; \textbf{gonzález-leonardo2022a?})
showing in some instances a ``doghnut effect'', with a decline in
population movements directed to cities and a rise in the inflow of
people to less dense surrounding areas (\textbf{ramani2021?}). Studies
demonstrated that professionals (\textbf{tønnessen2021?}) and
individuals with a high income (\textbf{haslag2021?}), those who are
potentially able to practice remote working, underpinned movements away
from large cities. In addition, research has shown that a large share of
counterurbanisation movements is headed to touristic locations with a
high concentration of second homes, suggesting that middle and high
class individuals may have been key actors in movements from cities to
less populated areas during the pandemic
(\textbf{gonzález-leonardo2022a?}). However, most studies suggested that
variations in internal population movements during the pandemic are
likely to be temporary and have not altered previous macro-structures of
human mobility across the rural-urban hierarchy (\textbf{rowe2023?}).

In Latin America and the Caribbean, a reduction in internal population
movements was documented during periods of high stringency
(\textbf{aromí2023?}), but there is no evidence on how human mobility
has changed across the rural-urban continuum. Anecdotal reports have
shown that movements away from cities seem to have accelerated in two
countries of the Global South, India (\textbf{irudayarajan2020?}) and
South Africa (\textbf{ginsburg2022?}), while urbanisation trends have
decelerated. However, lack of suitable up-to-date data has prevented us
from properly assessing the ``urban exodus'' hypothesis in Global South
countries, as well as the extent, spatial patterns, and durability of
potential changes on internal population movements. The evidence for
these two countries suggested that increasing counterurbanisation
movements during the pandemic in India and South Africa could be
underpinned by internal migrants residing in cities who returned to
their hometowns due to a rise in unemployment
(\textbf{irudayarajan2020?}; \textbf{ginsburg2022?}). However, recent
studies have shown that urban residents from wealthy neighborhoods were
more likely to move to rural areas than those from low income
neighborhoods during the first wave of the pandemic in Brazil, Colombia,
Mexico, Indonesia, Philippines and South Africa
(\textbf{lucchini2023?}).

It is known that human mobility within countries tends to vary in
systematic ways with development levels (\textbf{bell2015?}). In Latin
America and the Caribbean, 28.6\% of the population was living on less
than \$6.8 per day in 2020 (\textbf{WorldBank2023?}) and the informality
rate was about 53\% in 2016 (\textbf{OIT2018?}). Thus, we could expect
different pandemic impacts on the patterns of internal population
movements between countries in the Global North and the Global South.
The way in which COVID-19 unfolded in each country, with varying levels
of stringency measures, may have also produced different outcomes on
human mobility.

Traditionally, censuses have been used to analyse human mobility
patterns within Global South countries, which lack register data
(\textbf{bell2015?}; \textbf{bernard2017?}). Censuses, however, are
normally updated every 10 years, lacking the temporal granularity to
explore population movements over yearly, monthly or short-time
intervals.

Digital traces data from mobile phone applications, on the other hand,
provide a unique opportunity to capture internal population movements
with small spatial and temporal granularity in multiple countries
(\textbf{green2021?}; \textbf{rowe2021?}). Drawing on geographically
granular Meta-Facebook data from March 2020 to March 2022, we aim to
analyse how the pandemic has changed the patterns of human mobility
across the rural-urban hierarchy in four Latin American countries:
Argentina, Chile, Colombia and Mexico. We aim to address the following
research questions: 1) To what extent did people leave cities and move
to less populated areas in these countries? 2) Which are the spatial
patterns of potential changes on internal population movements? 3) How
have the patterns of human mobility changed over the course of the
pandemic? 4) Are potential changes likely to endure the pandemic?

@@

The COVID-19 pandemic led to an overall decline in human mobility within
countries (\textbf{nouvellet2021?}). Research has shown that internal
population movements across the rural-urban continuum decreased in the
United States (\textbf{ramani2021?}), several European countries, Japan
and Austria during 2020 (\textbf{rowe2023?}). Declines ranged from 2.5\%
in Spain to 8.5\% in Australia. The largest drops occurred during
periods of national lockdowns and the implementation of other stringency
measures, such as mobility restrictions, work and school closing, while
levels of internal population movements recovered or exceeded
pre-pandemic figures when lockdowns were lifted. Declines were partially
attributed to a reduction in involuntary migration during lockdowns, a
loss of labour market dynamism due to the pandemic, fewer people
changing jobs and increasing remote work (\textbf{perales2022?}).

COVID-19 has also altered patterns of internal population movements
across the rural-urban hierarchy (Rowe et al. 2022). In large metro
areas across the United States, population movements, as well as
households and businesses, shifted from central cities to suburbs and
exurbs during 2020 (\textbf{ramani2021?}). These authors label this
trend as a ``doughnut effect'', reflecting a decrease in population
inflows and activities in city centers and a growth in suburban rings.
They found a sizeable ``doghnut effect'' in large cities, a smaller
effect in medium-sized cities and almost no effect for small cities. A
similar trend was observed in Norway (\textbf{tønnessen2021?}), Germany
(\textbf{stawarz2022?}), Sweden (\textbf{vogiazides2022?}) and Japan
(\textbf{fielding2021?}; \textbf{kotsubo2022?}), where net migration
rates in big cities declined, while population movement to their suburbs
increased. In addition, this trend was accompanied by a deceleration of
urbanisation trends and unusual population gains in rural areas due to
an acceleration in counterurbanisation movements, mostly in rural
locations which are known to be holiday destinations with a high
concentration of second homes and are generally located in the close
proximity to big cities. An increase of counterurbanisation was the main
pandemic outcome on human mobility in Spain
(\textbf{gonzález-leonardo2022b?}; \textbf{gonzález-leonardo2022a?}),
the United Kingdom (Wang et al. 2022; Rowe et al. 2022) and Australia
(\textbf{perales2022?}), where there is no evidence of a ``doghnut
effect'', since substantial variations in suburbs were not found across
these countries.

However, studies suggest that variations are likely to be temporary
(\textbf{rowe2023?}). In Spain, the urbanisation movement returned to
pre-pandemic levels when the lockdown ended in mid-June 2020
(\textbf{gonzález-leonardo2022b?}) and unusually high levels of
counterurbanisation persisted over 2021, although the majority of
movements continued to occur between and within urban areas
(\textbf{gonzález-leonardo2022a?}). In Australia, the pandemic has
caused minor changes in spatial patterns of internal migration, and its
effects were minimal by the end of 2020 (\textbf{perales2022?}). In the
United Kingdom, mobility patterns returned to those registered prior to
the pandemic after the easing of non-pharmaceutical interventions (Rowe
et al. 2022; Wang et al. 2022). These findings suggest that COVID-19
generated shock waves leading to temporary changes in the patterns of
internal population movement across the rural-urban continuum, but it
has not significantly altered the prevalent macro-structures of
population movement within countries.

Despite some evidence in the Global North, lees is known about the
impact of COVID-19 on human mobility in the Global South countries. In
Latin America and the Caribbean, internal population movements declined
by 10\% during periods of severe stringent measures, from 16-19\% in
Bolivia, Ecuador and Argentina to less than 3\% in Paraguay and
Venezuela (\textbf{aromí2023?}). However, there is no evidence about
pandemic outcomes on human mobility across the rural-urban continuum. To
date, surveys carried out in India (\textbf{irudayarajan2020?}) and
South Africa (\textbf{ginsburg2022?}) have shown anecdotal evidence
suggesting that urbanisation trends declined during the pandemic, while
movements from cities to rural areas increased. Both studies pointed out
that a number of rural residents initiating a migration decreased during
2020 and some labourers in major cities seem to have returned to their
hometowns due to job losses, since stringency measures have extensively
impacted the functioning of the economy (\textbf{ghosh2020?}). Many
internal migrants residing in Indian cities lost their jobs and were
unable to afford basic necessities (\textbf{ILO2020?}). Collectively,
these findings suggest that vulnerable populations seem to have played a
role in increasing counterurbanisation movements in Global South
countries. However, recent work has shown that high-wealth groups
residing in cities across Mexico, Colombia, Brazil, Indonesia,
Philippines and South Africa were on average 159\% more likely to have
moved to rural area compared to those of low-wealth groups during early
stages of the pandemic (\textbf{lucchini2023?}), in line with the
evidencee found for Global North countries.

@@

Until 1980, rural to urban migration dominated internal population
movements in Latin America with significant levels of population
redistribution, especially during the rapid industrialisation process
between the 1950s and 1970s when a critical mass of individuals moved
from villages and towns to urban areas (\textbf{firebaugh1979?};
\textbf{lattes95?}; \textbf{sobrino12?}). Urban growth due to internal
mobility was primarily concentrated in large urban centres resulting in
a significant population imbalance between the chief cities and other
settlements (\textbf{PintoDaCunha02?}; \textbf{lattes2017?}). As a
consequence, Latin America shows high rates of urbanisation with more
than 81\% of the population living in cities, the highest figure after
North America, and about 46\% of urban residents settled in cities with
more than 1 million inhabitants (\textbf{UNpopulation19?}).

Currently, internal population movements amongst cities dominate the
migratory system in Latin America, while rural to urban flows are of
less importance (\textbf{bernard2017?}; \textbf{rodríguez-vignoli2018?};
\textbf{UNpopulation19?}). According to the 2010 census round, 80\% of
internal migrants moved between cities
(\textbf{rodríguez-vignoli2018?}). Contrary to the period preceding the
1980s, medium-sizes cities (those between 500K and 1 million
inhabitants) experienced the highest net migration rates, while large
cities (\textgreater1 million) presented balanced rates and small cities
(\textgreater500K) lost individuals by internal population movements
(\textbf{rodríguez-vignoli2018?}). Changing patterns in human mobility
within Latin American countries were driven by a decline in internal
migration since the 1980s crisis (\textbf{chávezgalindo2016?}) and
deconcentration dynamics in large urban centres, such as Mexico City
(\textbf{sobrino2006?}) or Santiago de Chile
(\textbf{gonzálezollino2006?}), where long distance inflows have
declined over time. At the same time, there has been an increasing
domestic and foreign investment in export-oriented activities or tourism
industries in certain middle-sized cities, fostering the geographic
dispersal of employment opportunities and increasing the attractiveness
of these cities (\textbf{Brea03?}; \textbf{Pérez-Campuzano13?};
\textbf{chávezgalindo2016?}).

Suburbanisation also represents an important type of internal population
movement and has increased over time due to the growing expansion of
Latin American cities (\textbf{graizbord2007?};
\textbf{chávezgalindo2016?}). These movements are underpinned by flows
of middle- and upper-class families moving away from central cities to
surrounding auto-segregated areas of wealthy individuals
(\textbf{borsdorf2003?}; \textbf{rodríguezvignoli2017?}). Low-income
populations also settle in specific areas within suburbs where the cost
of living is usually cheaper than in cities (\textbf{janoschka2002?};
\textbf{rodríguezvignoli2017?}). A large proportion of both wealthy
individuals and low-income residents commute daily to central cities,
mainly for work reasons (\textbf{chávezgalindo2016?}). Recently,
reurbanisation dynamics have also been observed due to gentrification
processes in specific central areas (\textbf{sobrino12?};
\textbf{chávezgalindo2016?}). In this paper, we analyse how the COVID-19
pandemic has changed patterns of internal population movements across
the rural-urban hierarchy in Mexico, Colombia, Perú, Chile and
Argentina.

@@

\begin{figure}[H]

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Maas19}
Maas, P., S. Iyer, A. Gros, W. Park, L. McGorman, C. Nayak, and P. A.
Dow. 2019. {``Facebook Disaster Maps: Aggregate Insights for Crisis
Response and Recovery.''} In \emph{16th International Conference on
Information Systems for Crisis Response and Management}, 836--47.

\bibitem[\citeproctext]{ref-rowe2022}
Rowe, Francisco, Alessia Calafiore, Daniel Arribas-Bel, Krasen
Samardzhiev, and Martin Fleischmann. 2022. {``Urban Exodus?
Understanding Human Mobility in Britain During the COVID-19 Pandemic
Using Facebook Data.''} \url{https://doi.org/10.48550/ARXIV.2206.03272}.

\bibitem[\citeproctext]{ref-wang2022}
Wang, Yikang, Chen Zhong, Qili Gao, and Carmen Cabrera-Arnau. 2022.
{``Understanding Internal Migration in the UK Before and During the
COVID-19 Pandemic Using Twitter Data.''} \emph{Urban Informatics} 1 (1).
\url{https://doi.org/10.1007/s44212-022-00018-w}.
